---
title: "Microplastics Bioindicators: Bight 23 Regional Monitoring"
author: "Leah Thornton Hampton"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
---
# Load libraries 

```{r, Packages}

library(tidyverse)
library(ggpubr)

```

# Import data

```{r, Data Import}

mp_sed_con <- read.csv(file = "Data/sed_mp_con.csv")

condition_scores <- read.csv(file = "Data/b23_conditionscores.csv")

benthic_abundance <- read.csv(file = "Data/B23 Post SDR benthic data.csv")

bight_infauna_unified <- read.csv(file = "Data/ed14 historical Bight infauna for unified DB v2.csv")

```

# MPs x all Benthic-indices
```{r, Benthic Indices}

mpdata_scores <- left_join(mp_sed_con, condition_scores) |> 
  group_by(stationid, stratumtype, index, score, condition, cond_class, n_g_sed_dw) |> 
  summarise(.groups = "drop") |> 
  drop_na(score, n_g_sed_dw) 

scores <- mpdata_scores |> 
  ggplot(aes(x = score, y = n_g_sed_dw))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(~index)+
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
  facet_wrap(~index, scales = "free") +
  labs(x = "score", y = "")

scores

cond <- mpdata_scores |> 
  ggplot(aes(x = condition, y = n_g_sed_dw))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(~index)+
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
  facet_wrap(~index, scales = "free") +
  labs(x = "condition", y = "")

cond

cond_class <- mpdata_scores |> 
  ggplot(aes(x = cond_class, y = n_g_sed_dw))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(~index)+
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
  facet_wrap(~index, scales = "free") +
  labs(x = "condition", y = "")

cond_class

```

# MPs x M-AMBI

```{r, M-AMBI}

ambi <- mpdata_scores |> 
  filter(index == "M-AMBI")

# 1. Normality: Shapiro-Wilk test (p > 0.05 suggests normality)
shapiro.test(ambi$score)
shapiro.test(ambi$n_g_sed_dw) 

# 2. Normality: Q-Q plots
par(mfrow = c(1, 2))
qqnorm(ambi$score, main = "Q-Q Plot: score")
qqline(ambi$score, col = "steelblue")
qqnorm(ambi$n_g_sed_dw, main = "Q-Q Plot: n_g_sed_dw")
qqline(ambi$n_g_sed_dw, col = "steelblue")
par(mfrow = c(1, 1))

# 3. Linearity: scatter plot with LOESS and linear smoother
ggplot(ambi, aes(x = score, y = n_g_sed_dw)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "steelblue", se = TRUE) +
  geom_smooth(method = "loess", color = "tomato", se = FALSE, linetype = "dashed") +
  labs(title = "Linearity check: score vs n_g_sed_dw (M-AMBI)",
       subtitle = "Blue = linear fit, Red dashed = LOESS")

# 4. Outliers: boxplots
par(mfrow = c(1, 2))
boxplot(ambi$score, main = "Outlier check: score")
boxplot(ambi$n_g_sed_dw, main = "Outlier check: n_g_sed_dw")
par(mfrow = c(1, 1))

# 5. Spearman's correlation test (all of the Pearson's assumptions violated)
cor.test(ambi$score, ambi$n_g_sed_dw, method = "spearman")

corr <- ambi |> 
  ggplot(aes(x = score, y = n_g_sed_dw))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
  facet_wrap(~index, scales = "free") +
  labs(x = "M-AMBI score", y = "")

corr
```

# Taxon Abundance Correlation

```{r}

#Join microplastics sediment concentration data with benthic taxon abundance

mpdata_abundance <- left_join(mp_sed_con, benthic_abundance, by = "stationid") |>
  select(-X, -lab.x, -lab.y, -sampledate, -replicate, -voucher, -personalvoucher, -qualifier, -taxonomist )

#Check where data is missing
mpdata_abundance |> 
  filter(is.na(abundance)) |> 
  distinct(stationid) |> 
  nrow()

#21 stations with mp data do not have microplastics data
#drop stations with no abundance data from analysis

mpdata_abundance <- mpdata_abundance |>
  drop_na(abundance)



#taxon-abundance-correlation

taxon_cor <- mpdata_abundance |>
  group_by(taxon) |>
  filter(n() >= 5) |>
  summarise(
    n_stations = n(),
    spearman_r = cor(abundance, n_g_sed_dw, method = "spearman", use = "complete.obs"),
    p_value = cor.test(abundance, n_g_sed_dw, method = "spearman")$p.value,
    .groups = "drop"
  ) |>
  arrange(p_value) |>
  as.data.frame()

taxon_sig_neg_cor <- taxon_cor |>
  filter(spearman_r < 0 & p_value < 0.05)

#taxon-correlation-plot

##As expected some species are positively correlated and some are negatively correlated with MP abundance. This on its own isn't very useful. Need to look if species traits or families are positively or negatively correlated

taxon_cor |>
  mutate(
    sig = case_when(
      p_value < 0.001 ~ "p < 0.001",
      p_value < 0.01  ~ "p < 0.01",
      p_value < 0.05  ~ "p < 0.05",
      TRUE            ~ "ns"
    ),
    sig = factor(sig, levels = c("p < 0.001", "p < 0.01", "p < 0.05", "ns")),
    taxon = reorder(taxon, spearman_r)
  ) |>
  ggplot(aes(x = spearman_r, y = taxon, fill = sig)) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c(
    "p < 0.001" = "#2166ac",
    "p < 0.01"  = "#74add1",
    "p < 0.05"  = "#abd9e9",
    "ns"        = "grey75"
  )) +
  labs(
    x = "Spearman r",
    y = NULL,
    fill = "Significance",
    title = "Taxon abundance vs. MP sediment concentration"
  ) +
  theme_minimal()

```

```{r}

##David mentioned that species traits (e.g., position, mobility, reproduction, feeding) vary at the family level. Join families to the abundance data and analyze

###NEED TO QA THIS -- NOT COMPLETE


#Check taxon overlap with historical Bight infauna
shared_taxa <- intersect(unique(mpdata_abundance$taxon), unique(bight_infauna_unified$taxon))
only_in_mpdata <- setdiff(unique(mpdata_abundance$taxon), unique(bight_infauna_unified$taxon))

cat("Taxa in both mpdata_abundance and bight_infauna_unified:", length(shared_taxa), "\n")
cat("Taxa only in mpdata (not in historical data):", length(only_in_mpdata), "\n")

#51 taxa in mpdata that don't match with taxa in unified data
##CHECK to see if this is real or if there are syntax issues

taxa_classification <- bight_infauna_unified |> 
  select(taxon, Phylum, Subphylum, Superclass, Class, Subclass, Infraclass, Superorder, Order, Suborder, Infraorder, Superfamily, Family) |> 
  distinct(taxon, Phylum, Subphylum, Superclass, Class, Subclass, Infraclass, Superorder, Order, Suborder, Infraorder, Superfamily, Family)

n_distinct(taxa_classification$taxon)    #2860
n_distinct(mpdata_abundance$taxon)  #787

mpdata_sp_class <- left_join(mpdata_abundance, taxa_classification, by = "taxon") ##add 


mpdata_sp_class <- mpdata_sp_class |>
  drop_na(Family)

family_abundance <- mpdata_sp_class |>
  group_by(stationid, Family, n_g_sed_dw, stratumtype) |>
  summarise(abundance = sum(abundance, na.rm = TRUE), .groups = "drop")



#family-abundance-correlation

family_cor <- family_abundance |>
  group_by(Family) |>
  filter(n() >= 5) |>
  summarise(
    n_stations = n(),
    spearman_r = cor(abundance, n_g_sed_dw, method = "spearman", use = "complete.obs"),
    p_value = cor.test(abundance, n_g_sed_dw, method = "spearman")$p.value,
    .groups = "drop"
  ) |>
  mutate(p_adj = p.adjust(p_value, method = "BH")) |>
  arrange(p_adj) |>
  as.data.frame()

family_sig_neg_cor <- family_cor |>
  filter(spearman_r < 0 & p_value < 0.05)

#family-correlation-plot

##As expected some families are positively correlated and some are negatively correlated with MP abundance.

family_cor |>
  mutate(
    sig = case_when(
      p_value < 0.001 ~ "p < 0.001",
      p_value < 0.01  ~ "p < 0.01",
      p_value < 0.05  ~ "p < 0.05",
      TRUE          ~ "ns"
    ),
    sig = factor(sig, levels = c("p < 0.001", "p < 0.01", "p < 0.05", "ns")),
    Family = reorder(Family, spearman_r)
  ) |>
  ggplot(aes(x = spearman_r, y = Family, fill = sig)) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c(
    "p < 0.001" = "#2166ac",
    "p < 0.01"  = "#74add1",
    "p < 0.05"  = "#abd9e9",
    "ns"        = "grey75"
  )) +
  labs(
    x = "Spearman r",
    y = NULL,
    fill = "Significance",
    title = "Family abundance vs. MP sediment concentration"
  ) +
  theme_minimal()


