---
title: "Toxicity Scoring"
format: html
---

This file assigns toxicity scores to species in the ToMEx database for the prioritization of candidate bioindicators for microplastics. Each species is scored based on 4 criteria within the toxicity category - sensitivity, dose response, level of biological organization, and weight of evidence.

```{r, packages and data}

library(tidyverse)

tox_dat <- readRDS(file = "Data/aoc_setup_tomex2.RDS")

```

```{r, data tidying}

```

# Sensitivity

0 – Effects observed only at concentrations greater than one order of magnitude of environmentally relevant concentrations
1 – Effects observed within an order of magnitude of environmentally relevant concentrations
2 – Effects observed at or below environmentally relevant concentrations

```{r, sensitivity}

```

# Dose Response

0 – No data sets available with at least 3 test concentrations in series or data sets with 3+ test concentrations do not demonstrate that response increases with increasing dose
1 – At least one data set available with 3-4 test concentrations, response increases with increasing dose
2 – At least one data set available with 5 test concentrations, response increases with increasing dose

```{r, dose response}

```

# Level of Biological Organization

0 – Effects only observed below the organism level
1 – Effects observed at the organism and population levels
2 – Effects observed at the organism and population levels pertaining to growth, development, reproduction, and survival

```{r, level of biological organization}

```

# Weight of Evidence

0 – Only 1 available study available or 2 studies with directly conflicting results
1 – At least 2 available studies and results generally agree
2 – More than 2 available studies and results generally agree

```{r, weight of evidence}

```

```{r, summary of species toxicity scores}

```

End of Analysis.