---
title: "Toxicity Scoring"
format: html
---

This file assigns toxicity scores to species in the ToMEx database for the prioritization of candidate bioindicators for microplastics. Each species is scored based on 4 criteria within the toxicity category - sensitivity, dose response, level of biological organization, and weight of evidence.
# Setup

```{r, packages and data}

library(tidyverse)
library(readxl)

tox_dat <- readRDS(file = "Data/aoc_setup_tomex2.RDS")

##Particle only data, Marine environment, particle aligned w/ alpha = marine surface water, upper particle length 50000 um
marine_particle_aligned_import <- read_excel("Data/ToMEx_Exploration2026-08-31_marine_particle_aligned.xlsx", sheet = 1)

```

# Data tidying

```{r, data tidying}

```

# Sensitivity

0 – Effects observed only at concentrations greater than one order of magnitude of environmentally relevant concentrations
1 – Effects observed within an order of magnitude of environmentally relevant concentrations
2 – Effects observed at or below environmentally relevant concentrations

```{r, sensitivity}


#What is an environmentally relevant concentration?
## Xhu et al. (2021) reported concentrations as high as 35.70 particles/Lin one-liter grab samples in the SF Bay(Table S15). Average grab sample concentrations were 4.1 +/- 6.3 particles/L. https://doi.org/10.1021/acsestwater.0c00292

## Singh et al. (2025) found concentrations ranged from 0.03 to 8.71 particles/L in San Pedro Bay water column (net sample collection). Mean surface concentrations were 2.06 particles per liter. Concentrations ranged from 500 to 3000 particles/kg dry weight in the sediment of the San Pedro Shelf (Van Veen grab)

## Environmental concentrations were "corrected" using the MaTCH tool at https://hannahhapich.shinyapps.io/match/. Data input file is "\Data\INPUT_match_tool.csv". Data output file is "\Data\OUTPUT_match_tool.csv"

#Importing corrected concentrations used on graph below:

corrected_concs <- read.csv(file = "Data/OUTPUT_match_tool.csv")

corrected_concs <- corrected_concs |> 
    mutate(corrected_particle_per_mL = corrected_concentration/1000)

#filter for only concentrations with effects
marine_particle_aligned <- marine_particle_aligned_import |> 
    filter(Effect == "Yes")

##Investigate level of biological organization, number of species, other parameters??. Add lines for mean and max concentrtions found in each study

n_species <- length(unique(marine_particle_aligned$Species))
n_groups <- length(unique(marine_particle_aligned$`Organism Group`))

ggplot(marine_particle_aligned, aes(x = "", y = `Plotted Dose Values`)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +
  geom_jitter(width = 0.1, alpha = 0.3) +
  geom_hline(data = corrected_concs, aes(yintercept = corrected_particle_per_mL), linetype = "dashed", color = "gray40") +
  geom_text(data = corrected_concs, aes(x = 1.3, y = corrected_particle_per_mL, label = sample_id), hjust = 0, size = 3, color = "gray20") +
  scale_y_log10() +
  scale_x_discrete(expand = expansion(add = c(0.5, 1))) +
  coord_cartesian(clip = "off") +
  labs(
    x = NULL,
    y = "Plotted Dose (particles/mL)",
    caption = paste0("Data for ", n_species, " species across ", n_groups, " organism groups")
  ) +
  theme_minimal() +
  theme(plot.caption = element_text(hjust = 0), plot.margin = margin(5.5, 5.5, 5.5, 5.5))

```

# Dose Response

0 – No data sets available with at least 3 test concentrations in series or data sets with 3+ test concentrations do not demonstrate that response increases with increasing dose
1 – At least one data set available with 3-4 test concentrations, response increases with increasing dose
2 – At least one data set available with 5 test concentrations, response increases with increasing dose

```{r, dose response}

```

# Level of Biological Organization

0 – Effects only observed below the organism level
1 – Effects observed at the organism and population levels
2 – Effects observed at the organism and population levels pertaining to growth, development, reproduction, and survival

```{r, level of biological organization}

```

# Weight of Evidence

0 – Only 1 available study available or 2 studies with directly conflicting results
1 – At least 2 available studies and results generally agree
2 – More than 2 available studies and results generally agree

```{r, weight of evidence}

```

```{r, summary of species toxicity scores}

```

End of Analysis.