---
title: "Toxicity Scoring"
format:
  html:
    code-fold: true
---

This file assigns toxicity scores to species in the ToMEx database for the prioritization of candidate bioindicators for microplastics. Each species is scored based on 4 criteria within the toxicity category - sensitivity, dose response, level of biological organization, and weight of evidence.
# Setup

```{r, packages and data}
#| output: false

library(tidyverse)
library(readxl)
library(ggrepel)
library(DT)

tox_dat <- readRDS(file = "Data/aoc_setup_tomex2.RDS")

##Particle only data, marine water, particle aligned w/ alpha = marine surface water, upper particle length 50000 um
marinewater_particle_aligned_import <- read_excel("Data/ToMEx_Exploration2026-08-31_marinewater_particle_aligned.xlsx", sheet = 1)
marinesed_particle_aligned_import <- read_excel("Data/ToMEx_Exploration2026-09-01_marinesed_particle_aligned.xlsx", sheet = 1)

```

# Data tidying

```{r, data tidying}

```

# Sensitivity

0 – Effects observed only at concentrations greater than one order of magnitude of environmentally relevant concentrations
1 – Effects observed within an order of magnitude of environmentally relevant concentrations
2 – Effects observed at or below environmentally relevant concentrations

```{r, sensitivity}
#| fig-width: 10
#| fig-height: 8
#| fig-dpi: 300

#What is an environmentally relevant concentration?
## Xhu et al. (2021) reported concentrations as high as 35.70 particles/Lin one-liter grab samples in the SF Bay(Table S15). Average grab sample concentrations were 4.1 +/- 6.3 particles/L. https://doi.org/10.1021/acsestwater.0c00292

## Singh et al. (2025) found concentrations ranged from 0.03 to 8.71 particles/L in San Pedro Bay water column (net sample collection). Mean surface concentrations were 2.06 particles per liter. Concentrations ranged from 500 to 3000 particles/kg dry weight in the sediment of the San Pedro Shelf (Van Veen grab)

## Environmental concentrations were "corrected" using the MaTCH tool at https://hannahhapich.shinyapps.io/match/. Data input file to MaTCH is aved as "\Data\INPUT_match_tool.csv". Data output file is "\Data\OUTPUT_match_tool.csv"

#Importing corrected concentrations used on graph below:

corr_envr_concs <- read.csv(file = "Data/OUTPUT_match_tool.csv")

corr_envr_concs <- corr_envr_concs |>
    mutate( display_id = case_when(
             sample_id == "Singh et al. 2025 (max surface concentration, manta trawl)" ~ "Singh et al. 2025 (Max conc.)",
             sample_id == "Singh et al. 2025 (mean surface concentration, manta trawl)" ~ "Singh et al. 2025 (Mean conc.)",
             sample_id == "Zhu et al. 2021 (max 1L grab conc.)" ~ "Zhu et al. 2021 (Max conc.)",
             sample_id == "Zhu et al. 2021 (mean 1L grab conc.)" ~ "Zhu et al. 2021 (Mean conc.)",
             sample_id == "Singh et al. 2025 (max sediment concentration, grab)" ~ "Singh et al. 2025 (Max sediment conc.)",
             sample_id == "Singh et al. 2025 (mean nearshore sediment concentration, grab)" ~ "Singh et al. 2025 (Mean sediment conc.)",
             sample_id == "Bight '23 (max sediment embayments)" ~ "Bight '23 (Max conc.)",
             sample_id == "Bight '23 (mean sediment embayments)" ~ "Bight '23 (Mean conc.)",
             TRUE ~ sample_id
           )) |> 
    mutate( unit = case_when(
      study_media == "marine surface" ~ "particles/L",
      study_media == "marine sediment" ~ "particles/kg",
      TRUE ~ study_media
    ))

#Water lines are converted to Particles/mL to match the marinewater Plotted Dose Values; sediment lines are already in Particles/kg sediment, matching the marinesed Plotted Dose Values, so no conversion is applied.
corrected_concs_water <- corr_envr_concs |>
  filter(study_media == "marine surface") |>
  mutate(corrected_particle_per_mL = corrected_concentration/1000)

corrected_concs_sed <- corr_envr_concs |> filter(study_media == "marine sediment")


#total data available, before filtering to significant effects only
n_species_total <- length(unique(marinewater_particle_aligned_import$Species))
n_groups_total <- length(unique(marinewater_particle_aligned_import$`Organism Group`))

#filter for only concentrations with effects
marinewater_particle_aligned <- marinewater_particle_aligned_import |>
    filter(Effect == "Yes")

##Investigate level of biological organization, number of species, other parameters??. Add lines for mean and max concentrtions found in each study

n_species <- length(unique(marinewater_particle_aligned$Species))
n_groups <- length(unique(marinewater_particle_aligned$`Organism Group`))

```

```{r, sensitivity scores}

## Selecting concentration in each matrix as "Environmentally relevant" threshold
envrconc_SW <- max(corrected_concs_water$corrected_particle_per_mL)
envrconc_SW_plusOM <- envrconc_SW*10

#for each species, take the lowest dose at which a significant effect was observed, then score against the environmental concentration thresholds
species_sensitivity <- marinewater_particle_aligned |>
    group_by(Species) |>
    summarize(`Organism Group` = first(`Organism Group`), min_dose = min(`Plotted Dose Values`, na.rm = TRUE)) |>
    mutate(Sensitivity = case_when(
      min_dose < envrconc_SW ~ 2,
      min_dose < envrconc_SW_plusOM ~ 1,
      TRUE ~ 0
    )) |>
    select(Species, `Organism Group`, Sensitivity) |>
    arrange(desc(Sensitivity))

```


# Dose Response

0 – No data sets available with at least 3 test concentrations in series or data sets with 3+ test concentrations do not demonstrate that response increases with increasing dose
1 – At least one data set available with 3-4 test concentrations, response increases with increasing dose
2 – At least one data set available with 5 test concentrations, response increases with increasing dose

```{r, dose response}

#Currently using "Effect == YES" as a proxy for increasing dose. Need to individually evaluate magnitude.
#`Number of Doses` is in addition to the control

#Magnitude of response isn't available in this data, so as a proxy: a data set "demonstrates" dose-response if, ordered by dose,
#Effect never reverts from "Yes" back to "No" at a higher dose (once an effect appears, it persists at higher doses).
species_dose_response <- marinewater_particle_aligned_import |>
    group_by(Species, DOI, Endpoint, `Broad Endpoint Category`, `Specific Endpoint Category`, `Exposure Route`, Polymer, Shape, `Size Category`, Sex, `Exposure Duration (days)`, `Number of Doses`, `Life Stage`, `In vitro/in vivo`) |>
    arrange(`Plotted Dose Values`, .by_group = TRUE) |>
    summarise(
      monotonic = {
        has_effect <- any(Effect == "Yes")
        if (!has_effect) {
          #effect never appears at any dose in this series, so no dose-response relationship to speak of
          FALSE
        } else {
          #index of the lowest dose where an effect first shows up (rows are already sorted by dose)
          first_yes_position <- min(which(Effect == "Yes"))
          #effect status at that dose and every higher dose after it
          effect_from_first_yes_onward <- Effect[first_yes_position:n()]
          #TRUE only if the effect stays "Yes" all the way up; FALSE if it ever reverts back to "No"
          all(effect_from_first_yes_onward == "Yes")
        }
      },
      .groups = "drop"
    ) |>
    mutate(dataset_score = case_when(
      monotonic & `Number of Doses` >= 5 ~ 2,
      monotonic & `Number of Doses` >= 3 ~ 1,
      TRUE ~ 0
    )) |>
    group_by(Species) |>
    summarize(`Dose Response` = max(dataset_score, na.rm = TRUE))

species_sensitivity <- species_sensitivity |>
    left_join(species_dose_response, by = "Species") |>
    arrange(desc(Sensitivity), desc(`Dose Response`))


```

# Level of Biological Organization

0 – Effects only observed below the organism level
1 – Effects observed at the organism and population levels
2 – Effects observed at the organism and population levels pertaining to growth, development, reproduction, and survival

```{r, level of biological organization}

#Score 2 requires an effect at the organism/population level AND that effect pertaining to growth, development, reproduction, or survival (mortality)
species_bio_org <- marinewater_particle_aligned |>
    group_by(Species) |>
    summarize(
      has_organism_pop_effect = any(`Level of Biological Organization` %in% c("Organism", "Population")),
      has_gdrs_effect = any(`Level of Biological Organization` %in% c("Organism", "Population") &
                                          `Specific Endpoint Category` %in% c("Mortality", "Growth", "Reproduction", "Development"))
    ) |>
    mutate(`Level of Biological Organization` = case_when(
      has_gdrs_effect ~ 2,
      has_organism_pop_effect ~ 1,
      TRUE ~ 0
    )) |>
    select(Species, `Level of Biological Organization`)

species_sensitivity <- species_sensitivity |>
    left_join(species_bio_org, by = "Species") |>
    arrange(desc(Sensitivity), desc(`Dose Response`))



```

# Weight of Evidence

0 – Only 1 available study available or 2 studies with directly conflicting results
1 – At least 2 available studies and results generally agree
2 – More than 2 available studies and results generally agree

```{r, weight of evidence}

```

# Summary toxicity scores

```{r, summary of species toxicity scores}

species_sensitivity <- species_sensitivity |>
    mutate(`Total Score` = Sensitivity + `Dose Response` + `Level of Biological Organization`) |>
    arrange(desc(`Total Score`))

DT::datatable(species_sensitivity)
```

# Data Visualization

```{r, Graph sensitivity by organism group}
#| fig-width: 10
#| fig-height: 8
#| fig-dpi: 300

#12 organism groups is too many to color distinctly and stay colorblind-safe; lump the 4 smallest into "Other"
marinewater_particle_aligned <- marinewater_particle_aligned |>
    mutate(organism_group_display = fct_lump_n(`Organism Group`, n = 8, other_level = "Other"))

organism_colors <- c(
  "Fish" = "#2a78d6",
  "Crustacea" = "#eb6834",
  "Mollusca" = "#1baf7a",
  "Algae" = "#eda100",
  "Cnidaria" = "#e87ba4",
  "Bacterium" = "#008300",
  "Rotifera" = "#4a3aa7",
  "Echinoderm" = "#e34948",
  "Other" = "#898781"
)

ggplot(marinewater_particle_aligned, aes(x = "", y = `Plotted Dose Values`)) +
  geom_boxplot(outlier.shape = NA, width = 1) +
  geom_jitter(aes(color = organism_group_display), width = 0.35, alpha = 0.6) +
  scale_color_manual(values = organism_colors, name = "Organism Group") +
  geom_hline(data = corrected_concs_water, aes(yintercept = corrected_particle_per_mL), linetype = "dashed", color = "gray40", alpha = 0.6) +
  geom_text(data = corrected_concs_water, aes(x = 1.7, y = corrected_particle_per_mL, label = display_id), hjust = 0, vjust = -0.25, size = 4, color = "gray20") +
  scale_y_log10() +
  scale_x_discrete(expand = expansion(add = c(0.75, 1.5))) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Marine Water",
    x = NULL,
    y = "Dose (particles/mL)",
    caption = paste0("Data available for ", n_species_total, " species across ", n_groups_total, " organism groups.\nData with significant effects are plotted, representing ", n_species, " species across ", n_groups, " organism groups.")
  ) +
  theme_minimal(base_size = 15) +
  theme(plot.caption = element_text(hjust = 0, size = rel(0.9)), plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        panel.background = element_blank(), plot.background = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        axis.ticks.y = element_line(color = "black"))

```

```{r, Graph sensitivity by level of biological organization}
#| fig-width: 10
#| fig-height: 8
#| fig-dpi: 300

marinewater_particle_aligned <- marinewater_particle_aligned |>
    mutate(bio_org_display = fct_collapse(`Level of Biological Organization`,
             `Subcell/Cell/Tissue` = c("Subcell", "Cell", "Tissue")))

bio_org_colors <- c(
  "Organism" = "#eda100",
  "Population" = "#2a78d6",
  "Subcell/Cell/Tissue" = "#898781"
)

bio_org_alphas <- c(
  "Organism" = 0.8,
  "Population" = 0.8,
  "Subcell/Cell/Tissue" = 0.2
)

ggplot(marinewater_particle_aligned, aes(x = "", y = `Plotted Dose Values`)) +
  geom_boxplot(outlier.shape = NA, width = 1) +
  geom_jitter(aes(color = bio_org_display, alpha = bio_org_display), width = 0.35) +
  scale_color_manual(values = bio_org_colors, name = str_wrap("Level of Biological Organization", width = 20), breaks = c("Organism", "Population", "Subcell/Cell/Tissue")) +
  scale_alpha_manual(values = bio_org_alphas, guide = "none") +
  geom_hline(data = corrected_concs_water, aes(yintercept = corrected_particle_per_mL), linetype = "dashed", color = "gray40", alpha = 0.6) +
  geom_text(data = corrected_concs_water, aes(x = 1.7, y = corrected_particle_per_mL, label = display_id), hjust = 0, vjust = -0.25, size = 4, color = "gray20") +
  scale_y_log10() +
  scale_x_discrete(expand = expansion(add = c(0.75, 1.5))) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Marine Water",
    x = NULL,
    y = "Dose (particles/mL)",
    caption = paste0("Data available for ", n_species_total, " species across ", n_groups_total, " organism groups.\nData with significant effects are plotted, representing ", n_species, " species across ", n_groups, " organism groups.")
  ) +
  theme_minimal(base_size = 15) +
  theme(plot.caption = element_text(hjust = 0, size = rel(0.9)), plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        panel.background = element_blank(), plot.background = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        axis.ticks.y = element_line(color = "black"))

```

```{r, Graph sensitivity sediment by organism group}
#| fig-width: 10
#| fig-height: 8
#| fig-dpi: 300

#total data available, before filtering to significant effects only
n_species_total_sed <- length(unique(marinesed_particle_aligned_import$Species))
n_groups_total_sed <- length(unique(marinesed_particle_aligned_import$`Organism Group`))

#filter for only concentrations with effects
marinesed_particle_aligned <- marinesed_particle_aligned_import |>
    filter(Effect == "Yes")

n_species_sed <- length(unique(marinesed_particle_aligned$Species))
n_groups_sed <- length(unique(marinesed_particle_aligned$`Organism Group`))

#only 2 organism groups present (Annelida, Mollusca); Mollusca kept the same color as the marinewater plot for consistency
sed_organism_colors <- c(
  "Annelida" = "#2a78d6",
  "Mollusca" = "#1baf7a"
)

ggplot(marinesed_particle_aligned, aes(x = "", y = `Plotted Dose Values`)) +
  geom_boxplot(outlier.shape = NA, width = 1) +
  geom_jitter(aes(color = `Organism Group`), width = 0.35, alpha = 0.6) +
  scale_color_manual(values = sed_organism_colors, name = "Organism Group") +
  geom_hline(data = corrected_concs_sed, aes(yintercept = corrected_concentration), linetype = "dashed", color = "gray40", alpha = 0.6) +
  geom_text_repel(data = corrected_concs_sed, aes(x = 1.5, y = corrected_concentration, label = display_id), hjust = 0, direction = "y", nudge_y = 0.2, segment.size = 0.3, min.segment.length = 0, size = 4, color = "gray20") +
  scale_y_log10() +
  scale_x_discrete(expand = expansion(add = c(0.75, 1.5))) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Marine Sediment",
    x = NULL,
    y = "Dose (particles/kg sediment)",
    caption = paste0("Data available for ", n_species_total_sed, " species across ", n_groups_total_sed, " organism groups.\nData with significant effects are plotted, representing ", n_species_sed, " species across ", n_groups_sed, " organism groups.")
  ) +
  theme_minimal(base_size = 15) +
  theme(plot.caption = element_text(hjust = 0, size = rel(0.9)), plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        panel.background = element_blank(), plot.background = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        axis.ticks.y = element_line(color = "black"))

```

```{r, Graph sensitivity sediment by level of biological organization}
#| fig-width: 10
#| fig-height: 8
#| fig-dpi: 300

marinesed_particle_aligned <- marinesed_particle_aligned |>
    mutate(bio_org_display = fct_collapse(`Level of Biological Organization`,
             `Subcell/Cell/Tissue` = intersect(c("Subcell", "Cell", "Tissue"), `Level of Biological Organization`)))

ggplot(marinesed_particle_aligned, aes(x = "", y = `Plotted Dose Values`)) +
  geom_boxplot(outlier.shape = NA, width = 1) +
  geom_jitter(aes(color = bio_org_display, alpha = bio_org_display), width = 0.35) +
  scale_color_manual(values = bio_org_colors, name = str_wrap("Level of Biological Organization", width = 20), breaks = c("Organism", "Population", "Subcell/Cell/Tissue")) +
  scale_alpha_manual(values = bio_org_alphas, guide = "none") +
  geom_hline(data = corrected_concs_sed, aes(yintercept = corrected_concentration), linetype = "dashed", color = "gray40", alpha = 0.6) +
  geom_text_repel(data = corrected_concs_sed, aes(x = 1.5, y = corrected_concentration, label = display_id), hjust = 0, direction = "y", nudge_y = 0.2, segment.size = 0.3, min.segment.length = 0, size = 4, color = "gray20") +
  scale_y_log10() +
  scale_x_discrete(expand = expansion(add = c(0.75, 1.5))) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Marine Sediment",
    x = NULL,
    y = "Dose (particles/kg sediment)",
    caption = paste0("Data available for ", n_species_total_sed, " species across ", n_groups_total_sed, " organism groups.\nData with significant effects are plotted, representing ", n_species_sed, " species across ", n_groups_sed, " organism groups.")
  ) +
  theme_minimal(base_size = 15) +
  theme(plot.caption = element_text(hjust = 0, size = rel(0.9)), plot.margin = margin(5.5, 5.5, 5.5, 5.5),
        panel.background = element_blank(), plot.background = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        axis.ticks.y = element_line(color = "black"))

```



End of Analysis.